\documentclass[11pt, a4paper]{article}

% --- Packages ---
\usepackage{amsmath, amssymb, amsthm}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage[style=authoryear, backend=biber]{biblatex}
\addbibresource{refs.bib}

\geometry{margin=2.5cm}

% --- Theorem environments ---
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\numberwithin{equation}{subsection}
\numberwithin{definition}{subsection}
\numberwithin{theorem}{subsection}
\numberwithin{remark}{subsection}

% --- Notation ---
\newcommand{\E}{\mathbb{E}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\dd}{\,\mathrm{d}}
\newcommand{\ii}{\mathrm{i}}   % imaginary unit

\title{\textbf{The Heston Stochastic Volatility Model}\\
       \large Semi-Analytical Pricing, Monte Carlo Simulation, and Market Calibration}
\author{Lucas Hutton}
\date{May--June 2026}

\begin{document}
\maketitle
\tableofcontents
\newpage

% ===========================================================================
\section{Introduction}
% ===========================================================================
% TODO (Phase 5 — write last):
%   - Motivate stochastic volatility: the empirical failure of Black-Scholes
%     (vol smile, vol surface).
%   - State what this paper does: derives, implements, and calibrates the
%     Heston (1993) model.
%   - Briefly outline the structure of the paper.
%   - Include a real SPX implied vol surface plot as the opening figure.


% ===========================================================================
\section{Mathematical Preliminaries}
% ===========================================================================
% This section develops the mathematical language needed for the Heston model.
% It should be concise -- background, not contribution.

\subsection{Probability Theory and Stochastic Processes}

The definitions and notation in this section follow \textcite{shreve2004}, Chapters 1--3.

First, the probability measure.

\begin{definition}[Probability Space]
Let \(\Omega\) be a non-empty set, \(\mathcal{F}\) be a \(\sigma\)-algebra of subsets of \(\Omega\), and \(\mathbb{P}\) be a probability measure over \(\mathcal{F}\). The triple \((\Omega, \mathcal{F},\mathbb{P})\) is called a \textup{probability space}.
\end{definition}

\begin{definition}[Almost Surely]
Let \((\Omega, \mathcal{F},\mathbb{P})\) be a probability space. If a set \(A \in \mathcal{F}\) satisfies \(\mathbb{P}(A) = 1\), we say that the event A occurs \textup{almost surely}.
\end{definition}

We next formalise the concept of a stochastic process, the mathematical object used to model the random evolution of quantities such as asset prices.

\begin{definition}[Filtration]
Let \(\Omega\) be a non-empty set. Let T be a fixed positive number, and assume that for each \(t \in [0, T]\), there is a \(\sigma\)-algebra \(\mathcal{F}(t)\). Assume further that if \(s \leq t\), then every set in \(\mathcal{F}(s)\) is also in \(\mathcal{F}(t)\). Then we call the collection of \(\sigma\)-algebras \(\mathcal{F}(t)\), \(0 \leq t \leq T\), a \textup{filtration}.
\end{definition}

\begin{definition}[Adapted Stochastic Process]
Let \(\Omega\) be a non-empty sample space equipped with a filtration \(\mathcal{F}(t)\), \(0 \leq t \leq T\). Let \(X(t)\) be a collection of random variables indexed by \(t \in [0, T]\). We say this collection of random variables is an \textup{adapted stochastic process} if, for each t, the random variable \(X(t)\) is \(\mathcal{F}(t)\)-measurable.
\end{definition}

In the context of continuous-time mathematical models, the flow of public information is interpreted as a filtration. The models adapt asset prices, portfolio processes, and wealth processes to this filtration.

\begin{definition}[Martingale]
Let \((\Omega, \mathcal{F},\mathbb{P})\) be a probability space, let T be a fixed positive number, and let \(\mathcal{F}(t)\), \(0 \leq t \leq T\), be a filtration of sub-\(\sigma\)-algebras of \(\mathcal{F}\). Consider an adapted stochastic process \(M(t)\), \(0 \leq t \leq T\). If
    
    \[
    \mathbb{E}[M(t) \mid \mathcal{F}(s)] = M(s) \text{ for all } 0 \leq s \leq t \leq T,
    \]

we say this process is a \textup{martingale}. It has no tendency to rise or fall.

\end{definition}

\subsection{Brownian Motion}

\begin{definition}[Brownian Motion]
Let \((\Omega, \mathcal{F},\mathbb{P})\) be a probability space. For each \(w \in \Omega\), suppose that there is a continuous function \(W(t)\) of \(t \geq 0\) that satisfies \(W(0) = 0\) and that depends on \(w\). Then, \(W(t)\), \(t \geq 0\), is a \textup{Brownian motion} if for all \(0 = t_0 < t_1 < \cdots < t_m\) the increments
    \begin{equation}
    W(t_1) - W(t_0),\; W(t_2) - W(t_1),\; \ldots,\; W(t_m) - W(t_{m-1})
    \end{equation}

    are independent and each of these increments is normally distributed with

    \begin{equation}
    \mathbb{E}[W(t_{i+1}) - W(t_{i})] = 0
    \end{equation}
    \begin{equation}
    \mathrm{Var}[W(t_{i+1}) - W(t_{i})] = t_{i+1} - t_i
    \end{equation}

\end{definition} 

We will now highlight two key properties of Brownian motion.

\begin{theorem}
Brownian motion is a martingale.
\end{theorem}

\begin{proof}
Let \(0 \leq s \leq t\) be given. Then
    \begin{align*}
    \mathbb{E}[W(t) \mid \mathcal{F}(s)] &= \mathbb{E}[W(t) - W(s) + W(s) \mid \mathcal{F}(s)] \\
                                         &= \mathbb{E}[W(t) - W(s) \mid \mathcal{F}(s)] + \mathbb{E}[W(s) \mid \mathcal{F}(s)] \\
                                         &= \mathbb{E}[W(t) - W(s)] + W(s) \\
                                         &= W(s).
    \end{align*}
The second equality is linearity of conditional expectation. For the third: \(W(t) - W(s)\) is independent of \(\mathcal{F}(s)\), so conditioning drops; \(W(s)\) is \(\mathcal{F}(s)\)-measurable, so \(\mathbb{E}[W(s)\mid\mathcal{F}(s)] = W(s)\). The last equality uses \(\mathbb{E}[W(t)-W(s)] = 0\). See \cite[p.~98]{shreve2004}.
\end{proof}


\begin{theorem}
Let W be a Brownian motion, and let \([W, W](T)\) denote the quadratic variation of W up to time T. Then \([W, W](T) = T\) for all \(T \geq 0\) almost surely.
\end{theorem}

\begin{proof}
See \cite[p.~102]{shreve2004}.
\end{proof}

Brownian motion is the natural model for the random component of asset prices: its continuous paths, stationary independent increments, and Gaussian distribution are consistent with observed log-returns over short time intervals. The quadratic variation result \([W,W](T) = T\) is of particular importance and demanded the development of It\^{o} calculus, as will be shown in the next section.

\subsection{The It\^{o}-Doeblin Formula}

We will now introduce the stochastic calculus developed by It\^{o}. The proofs in this section have been largely omitted due to their complexity with respect to the scope of this project.

To begin, we construct a mathematically rigorous representation of the value of a portfolio that results from trading assets in continuous time. 

\begin{definition}[Simple Process]
    Let T be a fixed positive number, and let \(\Delta(t)\) be an adapted stochastic process. Now let \(\Pi = \{t_0,t_1,\ldots,t_n\}\) be a partition of \([0, T]\). If \(\Delta(t)\) is constant in t on each subinterval \([t_j,t_{j+1})\), we say that \(\Delta(t)\) is a \textup{simple process}.
\end{definition}

\begin{definition}[It\^{o}'s Integral]
    Fix a positive number T. Let \(\Delta(t)\) be an adapted stochastic process, and \(W(t)\), \(t \geq 0\) a Brownian motion. \textup{It\^{o}'s integral} is given by

    \begin{equation}
        \int_{0}^{T} \Delta(t) \, \mathrm{d} W(t)
        \label{ito}
    \end{equation}

Which, for simple processes, is defined explicitly as
\begin{equation}
    I(t) = \sum_{j=0}^{k-1} \Delta(t_j)[W(t_{j+1}) - W(t_j)] + \Delta(t_k)[W(t) - W(t_k)]
\end{equation}

For the general case, we extend the integral as a limit of the integral of simple integrands in a fashion similar to that of the Lebesgue integral. (see \cite[p.~126]{shreve2004}).

\end{definition}

% Justify relevance?

\begin{theorem}
    The It\^{o} integral defined by \eqref{ito} is a martingale.
\end{theorem}

\begin{proof}
    See \cite[p.~128]{shreve2004}.
\end{proof}

The following definition allows us to extend the calculus to stochastic processes more general than Brownian motion. Almost all stochastic processes are It\^{o} processes.

\begin{definition}[It\^{o} Process]
    Let \(W(t)\), \(t \geq 0\), be a Brownian motion, and let \(\mathcal{F}(t)\), \(t \geq 0\), be an associated filtration. An \textup{It\^{o} process} is a stochastic process of the form

    \begin{equation}
        X(t) = X(0) + \int_{0}^{t}\Delta(u) \, \mathrm{d}W(u) + \int_{0}^{t}\Theta(u) \, \mathrm{d}u
        \label{ito process}
    \end{equation}

    where \(X(0)\) is non-random and \(\Delta(u)\) and \(\Theta(u)\) are adapted stochastic processes.
    
\end{definition}

\begin{theorem}
    The quadratic variation of the It\^{o} process \eqref{ito process} is

    \begin{equation}
        [X, X](t) = \int_0^t \Delta^2(u) \, \mathrm{d}u
    \end{equation}
    
\end{theorem}

\begin{proof}
    See \cite[p.~143]{shreve2004}
\end{proof}

\begin{remark}
    In particular, an Itô process with \(\Delta \not\equiv 0\) has non-zero quadratic variation, in contrast to differentiable functions, which have zero quadratic variation.
\end{remark}

The following theorem gives us a direct rule to differentiate expressions of the form \(f(W(t)\), where \(f(x)\) is a differentiable function and \(W(t)\) is a Brownian motion. The typical chain-rule fails here because the non-zero quadratic variation of the It\^{o} process introduces an additional term.

\begin{theorem}[It\^{o}-Doeblin Formula]
    Let \(X(t)\), \(t \geq 0\), be an It\^{o} process, and let \(f(t, x)\) be a function for which the partial derivatives \(f_t(t, x)\), \(f_x(t, x)\), and \(f_{xx}(t, x)\) are defined and continuous. Then, for every \(T \geq 0\),

    \begin{equation}
        f(T, X(T)) = f(0, X(0)) 
        + \int_0^T \left(f_t + \Theta(t) f_x + \frac{1}{2}\Delta^2(t) f_{xx}\right)\mathrm{d}t 
        + \int_0^T \Delta(t) f_x \, \mathrm{d}W(t)
        \label{ito formula}
    \end{equation}  
\end{theorem}

\begin{proof}
    See \cite[p.~146]{shreve2004}.
\end{proof}

\begin{remark}
    The formula given in \eqref{ito formula} is mathematically rigourous. However, in the derivation of stochastic differential equations, the theorem is typically given in differential notation:

    \begin{equation}
        \mathrm{d}f = \left(f_t + \Theta(t) f_x + \frac{1}{2}\Delta^2(t) f_{xx}\right)\mathrm{d}t 
        + \Delta(t) f_x \, \mathrm{d}W(t)
    \end{equation}
    
\end{remark}

\begin{remark}
    When \(\Delta(t) = 1\) and \(\Theta(t) = 0\), i.e. \(X(t) = W(t)\), we recover the equivalent formulation for a Brownian motion:

    \begin{equation}
        \mathrm{d}f = \left( f_t + \frac{1}{2}f_{xx} \right) \, \mathrm{d}t  + f_x \, \mathrm{d}W(t)
    \end{equation}
    
\end{remark}

In asset price modeling, we typically consider multiple variables. The Heston model in particular requires both asset price and instantaneous variance. For this reason, we extend It\^{o}-Doeblin for multiple processes.

\begin{theorem}[Two-dimensional It\^{o}-Doeblin formula]
    Let \(f(t, x, y)\) be a function whose partial derivatives \(f_t\), \(f_x\), \(f_y\), \(f_{xx}\), \(f_{xy}\), \(f_{yx}\), and \(f_{yy}\) are all defined and continuous. Let \(X(t)\) and \(Y(t)\) be It\^{o} processes. The two-dimensional \textup{It\^{o}-Doeblin formula in differential form} is
    \begin{equation}
        \mathrm{d}f(t, X, Y) = f_t \, \mathrm{d}t + f_x \, \mathrm{d}X + f_y \, \mathrm{d}Y + \frac{1}{2} f_{xx} \, \mathrm{d}X \, \mathrm{d}X + f_{xy} \, \mathrm{d}X \, \mathrm{d}Y + \frac{1}{2} f_{yy} \, \mathrm{d}Y \, \mathrm{d}Y
    \end{equation}
\end{theorem}

\begin{proof}
    See \cite[p.~167]{shreve2004}
\end{proof}

\subsection{Risk-Neutral Pricing and Girsanov's Theorem}
% TODO (Phase 2):
%   - The no-arbitrage condition and the existence of a risk-neutral measure Q.
%   - Girsanov's theorem: change of drift via change of measure.
%   - The fundamental pricing formula: V_t = e^{-r(T-t)} E^Q[payoff | F_t].
%   - Briefly state the Fundamental Theorem of Asset Pricing.


% ===========================================================================
\section{The Black--Scholes Model}
% ===========================================================================
% This section exists to motivate WHY we need stochastic volatility.
% End with the vol smile as an empirical contradiction of BS assumptions.

\subsection{Geometric Brownian Motion}
% TODO (Phase 1):
%   - GBM dynamics: dS_t = mu S_t dt + sigma S_t dW_t.
%   - Apply Ito's lemma to derive the log-normal distribution of S_T.

\subsection{The Black--Scholes Formula}
% TODO (Phase 1):
%   - Sketch the risk-neutral pricing derivation of the BS call formula.
%   - State the closed-form result; define d_1 and d_2.
%   - Note the single key input: constant volatility sigma.

\subsection{Implied Volatility and the Volatility Smile}
% TODO (Phase 1):
%   - Define implied volatility as the sigma that equates BS price to market price.
%   - Show a real implied vol surface (SPX data): smile/skew across strikes,
%     term structure across maturities.
%   - Conclude: constant sigma cannot produce a smile -- BS is structurally
%     wrong. This is the direct motivation for the Heston model.


% ===========================================================================
\section{The Heston Model}
% ===========================================================================

\subsection{Model Specification}
% TODO (Phase 2):
%   - State the two SDEs for S_t and v_t.
%   - Define the correlation structure: dW^S dW^v = rho dt.
%   - Explain intuitively what each equation captures.

\subsection{Parameters}
% TODO (Phase 2):
%   - Table of all five parameters: kappa, theta, sigma, rho, v_0.
%   - Explain the financial and statistical interpretation of each.
%   - Note typical calibrated ranges for equity indices.

\subsection{The Feller Condition}
% TODO (Phase 2):
%   - State: 2 kappa theta > sigma^2.
%   - Prove (or sketch the proof) that this ensures v_t > 0 almost surely.
%   - Discuss practical implications when it is violated in calibration.

\subsection{Risk-Neutral Dynamics}
% TODO (Phase 2):
%   - Apply Girsanov's theorem to move from physical to risk-neutral measure.
%   - State the risk-neutral SDEs used for pricing.
%   - Note: kappa and theta are understood as risk-neutral parameters when
%     calibrated directly to market prices.


% ===========================================================================
\section{The Characteristic Function}
% ===========================================================================

\subsection{Motivation and Affine Structure}
% TODO (Phase 3):
%   - Define the characteristic function phi(u) = E^Q[e^{iu log S_T} | F_t].
%   - Explain why the affine structure of the Heston model allows a
%     closed-form solution.

\subsection{Derivation via Riccati ODEs}
% TODO (Phase 3):
%   - Assume phi = exp(A(u,tau) + B(u,tau) v_t), substitute into the
%     Feynman-Kac PDE, separate terms.
%   - Show that A and B satisfy a pair of Riccati ODEs.
%   - Solve the ODEs to obtain the closed-form expressions for A and B.

\subsection{The Numerically Stable Form}
% TODO (Phase 3):
%   - State the original Heston (1993) form.
%   - Explain the branch-cut discontinuity causing numerical errors for
%     long maturities (Albrecher et al. 2007).
%   - State the corrected numerically stable form used in the implementation.


% ===========================================================================
\section{European Option Pricing via Fourier Methods}
% ===========================================================================

\subsection{Gil-Pelaez Fourier Inversion}
% TODO (Phase 3):
%   - Derive the call price as a Fourier integral over the characteristic
%     function (Gil-Pelaez inversion formula).
%   - Show how this reduces to a one-dimensional numerical integral.

\subsection{The Carr--Madan FFT Method}
% TODO (Phase 3):
%   - Introduce the dampening factor alpha to ensure square-integrability.
%   - Derive the modified pricing formula suitable for FFT evaluation.
%   - Explain the FFT discretisation: grid spacing in frequency domain,
%     resulting strike grid.
%   - Discuss the speed advantage: pricing across a full strike grid
%     simultaneously.


% ===========================================================================
\section{Monte Carlo Simulation}
% ===========================================================================

\subsection{Discretisation Scheme}
% TODO (Phase 2):
%   - Euler-Maruyama scheme for the joint (S_t, v_t) system.
%   - Log-Euler scheme for S_t to prevent negative prices.
%   - Simulation of correlated Brownian increments via Cholesky decomposition.

\subsection{Variance Process Treatment}
% TODO (Phase 2):
%   - Problem: Euler-Maruyama can produce negative variance values.
%   - Full truncation scheme: floor v_t at zero in the diffusion term.
%   - Brief discussion of alternative schemes (reflection, QE) and why full
%     truncation is used here.

\subsection{Convergence and Validation}
% TODO (Phase 3 / Phase 5):
%   - Sanity checks: put-call parity, agreement with Fourier pricer.
%   - Convergence of MC estimate as n_paths increases.
%   - Standard error as a measure of Monte Carlo uncertainty.


% ===========================================================================
\section{Calibration to Market Data}
% ===========================================================================

\subsection{Problem Formulation}
% TODO (Phase 4):
%   - Define the calibration objective: minimise MSE between Heston-implied
%     vols and market-implied vols across the observed surface.
%   - Explain the choice of implied vol space over price space (equal
%     weighting across strikes and maturities; scale-invariance).

\subsection{Data and Preprocessing}
% TODO (Phase 4):
%   - Data source: SPY/SPX options chain via yfinance.
%   - Filtering: non-zero bid, moneyness 0.85--1.15, 7--365 days to expiry.
%   - Extraction of market implied vols using the Black-Scholes inverter.

\subsection{Optimisation Strategy}
% TODO (Phase 4):
%   - Parameter bounds for each of the five parameters.
%   - Global search via differential evolution to avoid local minima.
%   - Local refinement via Nelder-Mead from the global solution.
%   - Discussion of identifiability: which parameters are well-determined
%     and which are degenerate.


% ===========================================================================
\section{Results and Analysis}
% ===========================================================================

\subsection{Calibrated Parameters}
% TODO (Phase 5):
%   - Table of calibrated parameter values on a specific date.
%   - Check Feller condition; interpret each parameter financially.

\subsection{Volatility Surface Fit}
% TODO (Phase 5):
%   - Side-by-side plots: market implied vol surface vs Heston fitted surface.
%   - RMSE in implied vol space.
%   - Discussion of where the model fits well (near ATM) and where it
%     struggles (deep wings, short maturities).

\subsection{Monte Carlo vs Fourier Pricing Comparison}
% TODO (Phase 5):
%   - Table comparing MC and Fourier prices across a range of strikes and
%     maturities.
%   - Confirm agreement within Monte Carlo standard errors.
%   - Compare computation times.


% ===========================================================================
\section{Conclusion}
% ===========================================================================
% TODO (Phase 5):
%   - Summarise what was built and what was found.
%   - Reflect on the limitations of the Heston model (cannot fit short-dated
%     wings perfectly; parameter instability over time).
%   - Suggest natural extensions: rough volatility, jumps (Bates model),
%     or multi-factor stochastic vol.


% ===========================================================================
\printbibliography
% ===========================================================================

\end{document}